CUBE: An \tcb{Information-optimized parallel} Cosmological $N$-body Algorithm

### 2017-11-29 ms_caf/ms_caf.tex
Paper: compared with CLASS power, added PM decorrelation

Relative to the optimized computation optimizations, $N$-body simulations use considerable amount of memory to store the \tcb{information of particles}. Their phase space coordinates $(x,y,z,v_x,v_y,v_z)$ are stored as at least 6 single-precision floating point numbers (total of 24 bytes). On the other hand, modern supercomputer systems use multi cores, many integrated cores (MIC) and even densely parallelized GPUs, bringing orders of magnitude higher computing power, whereas these architectures usually have limited memory allocation. Thus, these computation-light but memory-heavy applications, compared to matrix multiplication and decomposition calculations, are currently less suitable for fully usage of the computing power of modern supercomputers. For example, although native and offload modes of {\tt CUBEP3M} are able to run on the Intel Xeon-PHI MIC architectures, with the requirement of enough memory, TianNu simulation were done on TianHe-2 with only its CPUs -- 73\% of the total memory but only 13\% of the total computing power. \tcb{We investigate how most amount of information of particles can be deducted whereas still preserves the accuracy of $N$-body simulations, in order that the total memory usage for given $N$ is optimized.}

@@

First, by using different configurations -- different number of computing nodes, box sizes, particle resolutions, different number of tiles per node/image etc., we find that by using 2-byte integers for both positions and velocities ($n_\chi=n_\nu=2$, or x2v2) {\tt CUBE} gives exact results compared to {\tt CUBEP3M}. So, if memory usage is not a problem, one can always double the memory usage to get exact results, and the optimal memory limit of this case is 12 bpp, still much lower than traditional methods. Next, we focus on the accuracy of the other three cases -- x1v2, x2v1 and x1v1.

We list the names and configurations of simulations used in Table \ref{t.sim}, where $N_{\rm node}$, $L$, $z_i$, $z_{\rm ckpt}$, $N_p$ and $m_p$ are respectively the number of computing nodes used, length of the side of the box, initial redshift, redshift where we do checkpoint, total number of particles and particle mass. These configurations are run by {\tt CUBEP3M}, x2v2, x1v2, x2v1 and x1v1 versions of {\tt CUBE}. Using different number of tiles per image gives exact same results. We also list the configurations for TianNu and TianZero \citep{2017NatAs...1E.143Y,2017RAA....17...85E} simulations as a reference.

@@

\begin{figure*}
\centering
  \includegraphics[width=1.0\linewidth]{diff_power.pdf}
 \caption{Fractional deviations in the power spectrum $\Delta^2(k)/\Delta^2_0(k)-1$ (upper axes) and decorrelation $\xi$ (lower axes) induced by the integer-based algorithms in {\tt CUBE}. The four panels show the results of two redshifts in S2048S and S2048L: $\{ z=1, z=0 \} \times \{ L=400{\rm Mpc}/h, L=1200{\rm Mpc}/h \}$. The four solid lines show the results from \{x2v2,x1v2,x2v1,x1v1\}, whose colors are shown in the legends in the last panel. The orange dashed lines show the scale $k=0.2\ k_{\rm Nyquist}$, and smaller scales are beyond our interest.}
\label{fig.ccc}
\end{figure*}
\subsection{Power spectrum}

@@

In S2048S and S2048L are two simulations with $2048^3$ particles on small ($L=400\ {\rm Mpc}/h$) and large ($L=1200\ {\rm Mpc}/h$) box sizes. We compare their accuracy by their power spectra and their cross correlations with {\tt CUBEP3M} at two checkpoints $z=1$ and $z=0$. The physical scale of S2048S is designed such that the particle mass resolution $m_p$ is comparable to TianNu and TianZero simulations (their parameters are also listed in Table \ref{t.sim}) to study halo properties. On the other hand, S2048L focuses on larger structures, on which scale one can study weak gravitational lensing, baryonic acoustic oscillation (BAO) \citep{2005ApJ...633..560E} and its reconstruction \citep{2007ApJ...664..675E,2017ApJ...841L..29W} etc.

For each simulation the particles are firstly cloud-in-cell (CIC) interpolated onto the fine mesh grid of simulation, and from the density field $\rho$ we define the density contrast $\delta\equiv \rho/\left\langle \rho \right\rangle-1$. More generally we define the cross power spectrum $P_{\alpha\beta}(k)$ between two fields $\delta_\alpha$ and $\delta_\beta$ ($\delta_\alpha=\delta_\beta$ for auto power spectrum) in Fourier space as
\begin{equation}\label{eq.cps}
	\left\langle \delta_\alpha^\dagger({\bs k})\delta_\beta({\bs k}') \right\rangle=
    (2\pi)^3 P_{\alpha\beta}(k){\bs \delta}_{\rm 3D}({\bs k}-{\bs k}'),
\end{equation}
where ${\bs \delta}_{\rm 3D}$ is the three-dimensional Dirac delta function. In cosmology we usually consider the dimensionless power spectrum $\Delta^2_{\alpha\beta}(k)\equiv k^3 P_{\alpha\beta}(k)/(2\pi^2)$. The cross correlation coefficient is defined as
\begin{equation}\label{eq.ccc}
	\xi(k)\equiv P_{\alpha\beta}/\sqrt{P_{\alpha\alpha}P_{\beta\beta}}.
\end{equation}

We study the fractional power spectrum deviations $\Delta^2(k)/\Delta^2_0(k)-1$ (where $\Delta^2_0(k)$ denotes the {\tt CUBEP3M} power spectrum) and the cross correlation coefficient functions $\xi(k)$ between {\tt CUBE} and {\tt CUBEP3M}, and show the results in Figure \ref{fig.ccc}. For the four panels, the left two show S2048S and right two shows S2048L; the upper two show $z=1$ and lower two show $z=0$. \tcb{We label $k=0.2\ k_{\rm Nyquist}$ as dashed lines, where $k_{\rm Nyquist}$ is the scale of fine mesh grids, and the scale of average particle separations. From many simulations we observe that, on the scale $k=0.2\ k_{\rm Nyquist}$, the power spectrum is suppressed by $\sim 20\%$ compared to higher resolution simulations. This error is caused by the limited resolution of PM algorithms. Our scales of interest should be $k\lesssim 0.2\ k_{\rm Nyquist}$ where PM do not dominate the error, however in Figure \ref{fig.ccc} we still plot $k> 0.2\ k_{\rm Nyquist}$.}

The upper parts of the four panels show the fractional power spectrum deviations. \tcb{For both simulations and redshifts, the fractional power spectrum deviation is less than 1\% on scales of interest, where PM error do not dominate. If we just simply observe the scales where PM already contributes significant errors, there are differences between simulations.} For x2v2, it gives exact results as {\tt CUBEP3M} on all scales \tcb{(even including the scale where PM contribute significant error)} $\Delta^2_{\rm x2v2}(k)/\Delta^2_0(k)-1=0$ for both simulations and redshifts, shown by the four black straight lines. Similar to the results form Figure \ref{fig.dsp}, x1v2 (red) curves give better results than x2v1 (blue) and x1v1 (green), and x2v1 and x1v1 give nearly same results. Another trend is that higher redshifts and/or larger box sizes gives much better results. In the case of S2048S $z=0$ it shows more deviations near Nyquist frequency.

The lower parts of the four panels show the cross correlation coefficient function between {\tt CUBE} and {\tt CUBEP3M}. \tcb{For both redshifts of S2048L and $z=1$ of S2048S, the results show perfect cross correlation with {\tt CUBEP3M}, even for x1v1. For $z=0$ of S2048S, only x2v1 and x1v1 cases decorrelate from {\tt CUBEP3M}, on the scale of $k=0.2\ k_{\rm Nyquist}$, where the PM error is $\sim 20\%$, the decorrelation is only by $\sim 5\%$.}

>>>>

Relative to the optimized computation optimizations, $N$-body simulations use considerable amount of memory to store the \tcb{information of particles}. Their phase space coordinates $(x,y,z,v_x,v_y,v_z)$ are stored as at least 6 single-precision floating point numbers (total of 24 bytes). On the other hand, modern supercomputer systems use multi cores, many integrated cores (MIC) and even densely parallelized GPUs, bringing orders of magnitude higher computing power, whereas these architectures usually have limited memory allocation. Thus, these computation-light but memory-heavy applications, compared to matrix multiplication and decomposition calculations, are currently less suitable for fully usage of the computing power of modern supercomputers. For example, although native and offload modes of {\tt CUBEP3M} are able to run on the Intel Xeon-PHI MIC architectures, with the requirement of enough memory, TianNu simulation were done on TianHe-2 with only its CPUs -- 73\% of the total memory but only 13\% of the total computing power. \tcb{We investigate how most amount of information of particles can be deducted whereas it still preserves the accuracy of $N$-body simulations, in order that the total memory usage for given $N$ is optimized.}

@@

First, by using different configurations -- different number of computing nodes, box sizes, particle resolutions, different number of tiles per node/image etc., we find that by using 2-byte integers for both positions and velocities ($n_\chi=n_\nu=2$, or x2v2) {\tt CUBE} gives exact results compared to {\tt CUBEP3M}. So, if memory usage is not a problem, one can always double the memory usage to get exact results as {\tt CUBEP3M}, and the optimal memory limit of this case is 12 bpp, still much lower than traditional methods. Next, we focus on the accuracy of the other three cases -- x1v2, x2v1 and x1v1.

We list the names and configurations of simulations used in Table \ref{t.sim}, where $N_{\rm node}$, $L$, $z_i$, $z_{\rm ckpt}$, $N_p$ and $m_p$ are respectively the number of computing nodes used, length of the side of the box, initial redshift, redshift where we do checkpoint, total number of particles and particle mass. These configurations are run by {\tt CUBEP3M}, x2v2, x1v2, x2v1 and x1v1 versions of {\tt CUBE} with same initial seeds. Using different number of tiles per image gives exact same results. We also list the configurations for TianNu and TianZero \citep{2017NatAs...1E.143Y,2017RAA....17...85E} simulations as a reference.

@@

@@

\begin{figure*}
\centering
  \includegraphics[width=1.0\linewidth]{ccc.pdf}
 \caption{\tcb{Fractional deviations $\Delta^2(k)/\Delta^2_0(k)-1$ in the power spectra relative to nonlinear predictions $\Delta^2_0(k)$ (upper axes), and {\tt CUBE}'s decorrelations $\xi(k)$ relative to {\tt CUBEP3M} (lower axes) induced by the integer-based algorithms. The four panels show the results of two redshifts in S2048S and S2048L: $\{ z=1, z=0 \} \times \{ L=400{\rm Mpc}/h, L=1200{\rm Mpc}/h \}$. The four solid lines show the results from \{x2v2,x1v2,x2v1,x1v1\}. In lower panels, the dashed grey curves show the decorrelations given by the PM algorithm. The orange dashed lines show the scale $k=0.2\ k_{\rm Nyquist}$.}}
\label{fig.ccc}
\end{figure*}
\subsection{Power spectrum}

S2048S and S2048L are two simulations with $2048^3$ particles in small ($L=400\ {\rm Mpc}/h$) and large ($L=1200\ {\rm Mpc}/h$) box sizes. We compare their accuracy by their power spectra and their cross correlations with {\tt CUBEP3M} at two checkpoints $z=1$ and $z=0$. The physical scale of S2048S is designed such that the particle mass resolution $m_p$ is comparable to TianNu and TianZero simulations (their parameters are also listed in Table \ref{t.sim}) to study halo properties. On the other hand, S2048L focuses on larger structures, on which scale one can study weak gravitational lensing, baryonic acoustic oscillation (BAO) \citep{2005ApJ...633..560E} and its reconstruction \citep{2007ApJ...664..675E,2017ApJ...841L..29W} etc.

For each simulation, at redshifts $z=1$ and $z=0$, the particles are firstly cloud-in-cell (CIC) interpolated onto the fine mesh grid of simulation, and from the density field $\rho$ we define the density contrast $\delta\equiv \rho/\left\langle \rho \right\rangle-1$. More generally we define the cross power spectrum $P_{\alpha\beta}(k)$ between two fields $\delta_\alpha$ and $\delta_\beta$ ($\delta_\alpha=\delta_\beta$ for auto power spectrum) in Fourier space as
\begin{equation}\label{eq.cps}
	\left\langle \delta_\alpha^\dagger({\bs k})\delta_\beta({\bs k}') \right\rangle=
    (2\pi)^3 P_{\alpha\beta}(k){\bs \delta}_{\rm 3D}({\bs k}-{\bs k}'),
\end{equation}
where ${\bs \delta}_{\rm 3D}$ is the three-dimensional Dirac delta function. In cosmology we usually consider the dimensionless power spectrum $\Delta^2_{\alpha\beta}(k)\equiv k^3 P_{\alpha\beta}(k)/(2\pi^2)$. The cross correlation coefficient is defined as
\begin{equation}\label{eq.ccc}
	\xi(k)\equiv P_{\alpha\beta}/\sqrt{P_{\alpha\alpha}P_{\beta\beta}}.
\end{equation}

\tcb{We compute the nonlinear prediction of the power spectrum by CLASS \citep{2011JCAP...07..034B}, denoted $\Delta^2_0(k)$, at $z=0,1$. Then we study the fractional power spectrum deviations $\Delta^2(k)/\Delta^2_0(k)-1$. In the four panels of figure \ref{fig.ccc} we denote the box sizes of the simulation and redshifts at the upper left corners. The upper parts of these four panels show $\Delta^2(k)/\Delta^2_0(k)-1$, and the four curves of different colors show the results of x2v2, x1v2, x2v1 and x1v1. They almost overlap with each other, and in general agree with the nonlinear predictions.}

\tcb{We label $k=0.2k_{\rm Nyquist}$ as vertical dashed lines, where $k_{\rm Nyquist}$ is the scale of fine mesh grids, and the scale of average particle separations. On this scale, the power spectra are offset from nonlinear predictions by at least 20\%. This error is from the PM algorithm and one has to increase the resolution of the simulation to correct these offsets. We do not plot {\tt CUBEP3M} because we examined that {\tt CUBEP3M} and x2v2 of {\tt CUBE} produce same results. Thus, the differences between {\tt CUBEP3M} and different integer formats of {\tt CUBE} are negligible compared to the error of the PM algorithm.}

\tcb{In the lower parts of these four panels study the cross correlations. This time we in turn compare everything with {\tt CUBEP3M}, to emphasize the decorrelation by different integer formats. Firstly, x2v2 is perfectly correlated and cross correlation of x1v2 is nearly 100\% in all cases. In order to quantify the PM error in terms of decorrelation, we run x2v2 simulations with same initial seeds, but 8 times more particles, and measure the decorrelation caused by coarser resolutions. These results are dashed curves (labeled ``PM'') in the lower parts of these four panels. We conclude that cross correlation of x1v1, in all cases, is not worse than the PM errors.}
